Time and space specifications for stencils

### 2016-06-15 paper.tex
updated EBNF grammar with approximation info
grammar tweaks
corrected the comment syntax
Adjust specification rule & intro term, nonterm macros.
Adjust specification rule to be regionDec | specDec.
Introduce \term and \nonterm macros to replace \texttt and \textit.
fixed multiline syntax in fig1
improving intro

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}
\usepackage{amsthm}

@@

% Syntax macros
\newcommand{\stenFwd}[2]{\texttt{forward}, \texttt{depth}=#1,
  \texttt{dim}=#2}
\newcommand{\stenBwd}[2]{\texttt{backward}, \texttt{depth}=#1,
  \texttt{dim}=#2}
\newcommand{\stenCen}[2]{\texttt{backward}, \texttt{depth}=#1,
  \texttt{dim}=#2}

@@

\emph{Stencils} are a ubiquitous programming pattern in many numerical
applications. Informally, a stencil computation computes
an array, where the value at each index $i$ of this array is
calculated from a set of neighbouring values to $i$ in some input
array(s), \eg{}, the Game of Life, convolutions in image processing,
approximations to differential equations. For example, the following
psuedo-code iteratively applies the discrete Laplace
operator (an approximation to a derivative):
%
\begin{ExmVerbatim}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   a = b
\end{ExmVerbatim}
%
Line 3 is the core of the stencil computation, calculating
the value at \texttt{b[i]} from a neighbourhood of elements about
\texttt{a[i]}. Line 4 swaps
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration.

%This stencil computation exhibits statically decideable
%spatial and temporal relationships between $\texttt{a}$ and
%$\texttt{b}$.
In this simple example, the spatial and temporal relationship
between \texttt{a} and \texttt{b} is easy to understand. However,
more complex stencil computations can be much harder to understand
and subsequently more prone to error. For example, Figure~\ref{ref:navier-stokes-fragment}
shows a couple of lines from a Navier-Stokes fluid simulator in
which two arrays are read from with different data access patterns,
across two dimensions. The interaction is much harder to understand,
with the potential for the developer to accidentally introduce an
error, for example writing $\texttt{(i-1,j)}$ instead of
$\texttt{(i+1,j)}$.

We introduce a simple specification for stencil comptuations
that abstracts the fine grained detail of their access patterns.
In practise, most stencil computations have a regular shape that
can be described simply and abstractly. In the case of our
first Laplace example, our analysis tool infers the following specification:
%
\begin{SpecVerbatim}
%=  stencil centered, depth=1 dims=1 :: a
%=  stencil dependency a, mutual    :: b
\end{SpecVerbatim}
%
The first line explains that \texttt{a} is read from
with a symmetrical stencil pattern (``centered'') to a depth
of one in each direction in its first (and only) dimension.
The second line explains the temporal relationship between
\texttt{b} and \texttt{a}: that the previous time step for \texttt{b}
is actually provided by \texttt{a}. The dimension \texttt{t} is a
reserved dimension identifiers for temporal specifications. In
the case of the Navier-Stokes example, its inferred stencil is shown
in Figure~\ref{ref:navier-stokes-fragment}(b).

@@

\begin{SpecVerbatim}[xleftmargin=0.3cm]
%=   stencil (centered, depth=1, dim=1)
%=         + (centered, depth=1, dim=2) :: u

%=   stencil (forward, depth=1, dim=1)
%=         + (backward, depth=1, dim=1) :: v
\end{SpecVerbatim}
(b). Inferred stencil specification from CamFort
\caption{Fragment of Navier-Stokes fluid simulator and its specification}
\label{ref:navier-stokes-fragment}
\end{figure}

@@

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\textit{specification} ::= & [\textit{regionDec} \; \{\texttt{$\backslash$n} \;\;
                             \textit{regionDec} \}] \;
                             \textit{specDec} \\
\textit{specDec} ::= & \texttt{stencil} \; \textit{spec} \mid \epsilon
  \\
\textit{regionDec} ::= &  \texttt{region} \; \textit{rvar} \; \texttt{=} \;
                         \textit{region}\\
\textit{spec} ::= & \textit{spatial} \mid \textit{temporal}   \\[1em]
\textit{spatial} ::= & [\textit{mod} \; \{, \textit{mod} \}] \; \textit{region} \\
\textit{mod} ::= &
 \texttt{read-once} \\
\mid \; & \texttt{reflexive} \; \texttt{dims=}\dimTy \{\texttt{,} \; \dimTy\} \\
\mid \; & \texttt{irreflexive} \; \texttt{dims=}\dimTy \{\texttt{,} \;
          \dimTy\} \\[0.1em]
\textit{region} ::=
     \; & \texttt{forward} \; [,\texttt{depth=}\mathbb{N}] \; , \texttt{dim=}\dimTy  \\
\mid \; & \texttt{backward} \; [,\texttt{depth=}\mathbb{N}] \; , \texttt{dim=}\dimTy \\
\mid \; & \texttt{centered} \; [,\texttt{depth=}\mathbb{N}] \; , \texttt{dim=}\dimTy \\
\mid \; & \textit{region} \; \texttt{+} \; \textit{region} \\
\mid \; & \textit{region} \; \texttt{*} \; \textit{region} \\
\mid \; & \textit{rvar}  \\[0.5em]
\textit{temporal} ::= \; & \texttt{dependency} \; (v \; \{ , v \})
  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\textit{rvar} ::= \; & [\text{\texttt{a}-\texttt{z}$\,$\texttt{A}-\texttt{Z}$\,$\texttt{0}-\texttt{9}}]+
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

@@

@@

\begin{ExmVerbatim}
b(x) = a(x) + r*(a(x-1) - 2*a(x) + a(x+1))
\end{ExmVerbatim}
%
For this statement, the inference provides the specification:
%
\begin{SpecVerbatim}
%=  stencil centered, depth=1, dim=0 :: a
\end{SpecVerbatim}

@@

\paragraph{Proof}
\begin{itemize}
\item \textsc{coalesceF}
\begin{align*}
\llbracket & \texttt{v} : \texttt{fwd} \; \texttt{dims}=ds \; \texttt{depth}=n; \,
  \texttt{v} :  \texttt{fwd} \; \texttt{dims}=ds \;
  \texttt{depth}=n+1)\rrbracket \\
& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \relix, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq (n+1) \} \; \cup \; \{ ix^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq n \}\\
& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \relix, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ((ix \, d \leq (n+1)) \vee (ix \,
  d \leq n)) \} \\
& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \relix, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq (n+1) \} \\
& = \interp{\texttt{v} : \texttt{fwd} \; \texttt{dims}=ds \;
  \texttt{depth}=n + 1} \qquad \Box
\end{align*}
\end{itemize}

>>>>

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}
\usepackage{amsthm}
\usepackage{tikz}

@@

% Syntax macros
\newcommand{\nonterm}[1]{\textit{#1}}
\newcommand{\term}[1]{\texttt{#1}}

\newcommand{\stenFwd}[2]{\texttt{forward} \, (\texttt{depth=}#1,
  \texttt{dim=}#2)}
\newcommand{\stenBwd}[2]{\texttt{backward} \, (\texttt{depth=}#1,
  \texttt{dim=}#2)}
\newcommand{\stenCen}[2]{\texttt{centered} \, (\texttt{depth=}#1,
  \texttt{dim=}#2)}

@@

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation computes an array value, where the value at each index $i$ of
this array is calculated from a \emph{neighbourhood} of values around $i$ in
some input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following iteratively computes the one-dimensional discrete Laplace
transform (an approximation to a derivative): 
%
\begin{ExmVerbatim}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   a = b
\end{ExmVerbatim}
%
Line 3 is the core of the stencil computation, calculating
the value at \texttt{b(i)} from a neighbourhood of elements about
\texttt{a(i)}. Line 4 swaps
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration. (Note, we use the parentheses syntax
\texttt{( )} for
array subscripts rather than the bracket syntax \texttt{[ ]}.

%This stencil computation exhibits statically decideable
%spatial and temporal relationships between $\texttt{a}$ and
%$\texttt{b}$.
In this simple example, the dependencies between between \texttt{a}
and \texttt{b} form a simple spatial relationship which is easily
understood. This spatial relationship determines other aspects of the
program and its efficient implementation: how much ``boundary'' is
needed for the array, the most cache-efficient layout in memory,
the partitioning shape for parallel implementations.

More complex stencil computations can be much harder to understand and
subsequently more prone to error. For example,
Figure~\ref{ref:navier-stokes-fragment} shows three lines from a
Navier-Stokes fluid simulator in which two arrays are read from with
different data access patterns, across two dimensions. The interaction
is much harder to understand, with the potential for the developer to
accidentally introduce an error via simple textual mistakes, for
example writing $\texttt{(i-1,j)}$ instead of $\texttt{(i+1,j)}$.

In this work, we introduce a simple specification language for the
spatial and temporal properties of stencil comptuations. The
specifications abstract over the fine grained detail of stencil access
patterns.  In practise, most stencil computations have a regular shape
that can be described simply and abstractly, with a small set of
coarse-grained descriptions. In the case of our first Laplace example,
our inference procedure provides the specification:
%
\begin{SpecVerbatim}
!=  stencil centered (depth=1, dim=1) :: a
!=  stencil dependency a, mutual       :: b
\end{SpecVerbatim}
%
The first line explains that \texttt{a} is read from with a
symmetrical stencil pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.  The second line explains
the temporal relationship between \texttt{b} and \texttt{a}: that
previous time step for \texttt{b} is actually provided by \texttt{a},
and vice versa. This is explained as a mutual dependence between
\texttt{a} and \texttt{b}. In the case of the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Figure~\ref{ref:navier-stokes-fragment}(b). The 
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. These can drawn pictorially as follows where the horizontal
dimensions is \text{dim=1} and the vertical is \texttt{dim=2}:

\begin{center}
\begin{tikzpicture}
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}

@@

\begin{SpecVerbatim}[xleftmargin=0.3cm]
!=   stencil centered(depth=1, dim=1)
           + centered(depth=1, dim=2) :: u

!=   stencil forward(depth=1, dim=1)
           * backward(depth=1, dim=2) :: v
\end{SpecVerbatim}
(b). Inferred stencil specification from CamFort
\caption{Fragment of Navier-Stokes fluid simulator and its specification}
\label{ref:navier-stokes-fragment}
\end{figure}

@@

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec}
  \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\
\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}   \\[1em]
\nonterm{spatial} ::= & [\nonterm{mod}] \;
                       [\nonterm{approxMod},] \; [\nonterm{region}] \\
\nonterm{mod} ::= & [\nonterm{modR},] \; [\nonterm{modRefl},] \; [\nonterm{modIrefl},] \\
\nonterm{modR} ::= &
 \term{readOnce} \\
\nonterm{modRefl} ::= & \term{reflexive}, \; \term{dims=}\dimTy \{\term{,} \; \dimTy\} \\
\nonterm{modIrrefl} ::= & \term{irreflexive}, \; \term{dims=}\dimTy \{\term{,}
          \dimTy\}\\
\nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \term{forward}, \term{depth=}\mathbb{N}, \term{dim=}\dimTy  \\
\mid \; & \term{backward}, \term{depth=}\mathbb{N}, \term{dim=}\dimTy \\
\mid \; & \term{centered}, \term{depth=}\mathbb{N}, \term{dim=}\dimTy \\
\mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
\mid \; & \nonterm{region} \; \term{*} \; \nonterm{region} \\
\mid \; & \nonterm{rvar}  \\[0.5em]
\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \})
  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

@@

% Implicit vs explicit differentiation methods
% (forward implicit needs 'solver')
% backward - Euler
% runge-kutta 2nd-order

@@

\begin{ExmVerbatim}
b(x) = a(x) + r*(a(x-1) - 2*a(x) + a(x+1))
\end{ExmVerbatim}
%
For this statement, the inference provides the specification:
%
\begin{SpecVerbatim}
!=  stencil centered, depth=1, dim=0 :: a
\end{SpecVerbatim}

@@

\paragraph{Proof}
\begin{itemize}
\item \textsc{coalesceF}
\begin{align*}
\llbracket & \texttt{v} : \texttt{fwd} \; \texttt{dims}=ds \; \texttt{depth}=n; \,
  \texttt{v} :  \texttt{fwd} \; \texttt{dims}=ds \;
  \texttt{depth}=n+1)\rrbracket \\
& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \relix, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq (n+1) \} \; \cup \; \{ ix^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq n \}\\
& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \relix, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ((ix \, d \leq (n+1)) \vee (ix \,
  d \leq n)) \} \\
& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \relix, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq (n+1) \} \\
& = \interp{\texttt{v} : \texttt{fwd} \; \texttt{dim}=ds \;
  \texttt{depth}=n + 1} \qquad \Box
\end{align*}
\end{itemize}
